---
title: "qatar_final_assignment"
author: "Boglarka Ludmany"
date: "2026-05-12"
output: html_document
---

`echo = TRUE` - the code chunks are visible
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r}
library(tidyverse)
library(dplyr)
library(tidytuesdayR)
library(readr)
library(stats)
library(ggplot2)
library(magrittr)
```

# 1. Importing data - directly from GitHub
```{r}
qatarcars <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-12-09/qatarcars.csv")
```

# Questions to explore from the tidytuesday GitHub repository
- What's the distribution of price? (there are some really expensive cars here!)
- What's the relationship between (logged) price and performance?
- Are there patterns across cars from different countries? Do some countries make more expensive cars? More electric cars?
- What's the relationship between car dimensions and seating or trunk volume?

# 2. EDA
## 2.1. Exploring data structure, NA values 
```{r}
view(qatarcars)
str(qatarcars) ## 105 rows altogether, data types are mainly double and character 

colSums(is.na(qatarcars)) ## NA values only in the economy variable

economy_na <- qatarcars[!complete.cases(qatarcars),] ## investigating the reason behind NA values in the economy variable 
economy_na ## electric cars cannot be evaluated by their fuel economy
```

Value ranges with `summary`
```{r}
summary(qatarcars)
```
Unique values of `origin` and `make`
```{r}
origin_values <- tibble(unique(qatarcars$origin))
origin_values

brand_values <- tibble(unique(qatarcars$make))
brand_values
```


## 2.2. Visualizations, Distributions
### A. Univariate distributions
#### Horsepower
```{r}
range(qatarcars$horsepower)
horsepower_plot <- ggplot(qatarcars, aes(x = horsepower)) +
  geom_histogram(binwidth = 20) +
  scale_y_continuous(limits = c(0, 15),
                   breaks = waiver())

horsepower_plot
```
The distribution of horsepower is right-skewed: most cars are concentrated in the lower range, while a few high-performance outliers can also be observed. 

#### Type of the car (`type`)
```{r}
unique(qatarcars$type)

ggplot(qatarcars, aes(x = type)) +
  geom_histogram(stat = "count", 
                 color = "black", 
                 fill = "grey") +
  scale_y_continuous(limits = c(0, 80)) +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)), 
    vjust = -1
  )

## extracting the distribution of the types of cars from the original dataset  
type_sum <- qatarcars %>% 
  count(type, name = "count")
```
With regards to the distribution of car types, majority of the cars are SUVs (52.381%), 26.667% of all cars are Sedans, 13.333% are Coupes, and 7.762% are Hatchbacks. 

#### Type of the engine (`enginetype`)
```{r}
unique(qatarcars$enginetype)

enginetype_counts <- qatarcars %>% 
  count(enginetype, name = "count") %>%
  mutate(percent = round(count / sum(count) * 100, 1)) 

ggplot(enginetype_counts, aes(x = enginetype, y = count)) +
  geom_bar(stat = "identity", color = "black", fill = "grey") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  geom_text(
    aes(label = count),
    vjust = -1
  )
```
Majority of the cars are fueled by petrol (77.1%), hybrid (13.3%) and electric cars (9.5%) are present in smaller extent. 

#### Origin (`origin`)
```{r}
unique(qatarcars$origin)

origin_counts <- qatarcars %>% 
  count(origin, name = "count") %>%
  mutate(percent = round(count / sum(count) * 100, 1)) 

## this plot visualizes the percentages of each country of origin
ggplot(origin_counts, aes(x = origin, y = percent)) +
  geom_bar(stat = "identity", color = "black", fill = "grey") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  geom_text(
    aes(label = percent), 
    vjust = -1
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The majority of the cars originated from Japan (27.6%), followed by Germany (19%) and China (17.1%). A smaller proportion comes from the UK (10.5%), South Korea (9.5%), the USA (8.6%), Italy (4.8%), and Sweden (2.9%).

#### Car brands
```{r}
qatarcars %>% summarise(n_brands = n_distinct(make))

brands_counts <- qatarcars %>% 
  count(make, name = "count") %>%
  mutate(percent = round(count / sum(count) * 100, 1)) 

ggplot(brands_counts, aes(x = make, y = count)) +
  geom_bar(stat = "identity", color = "black", fill = "grey") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  geom_text(
    aes(label = count), 
    vjust = -1
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The dataset includes 31 distinct car brands, with Toyota being the most represented (_n_ = 10), followed by Kia (_n_ = 6), BMW (_n_ = 5), Mercedes (_n_ = 5), and MG (_n_ = 5). The majority of the brands appear only 3-4 times, three brands (Jaguar, Lotus, and Subaru) have only one observation. This distribution can be interpreted as highly uneven, indicating that brand-level comparisons should be interpreted with caution. 

### B. Multivariate Distributions
To explore the relationship between engine performance and fuel consumption, a scatterplot of horsepower versus fuel consumption (L/100 km), by engine type is presented.
```{r}
economy_data <- qatarcars %>% 
  filter(enginetype != "Electric") %>% ## fuel economy cannot be interpreted in the case of electric cars 
  select(enginetype, horsepower, economy)

ggplot(economy_data, aes(x = horsepower, y = economy, fill = enginetype)) +
  geom_point(aes(colour = enginetype)) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Relationship between horsepower and fuel economy, \nby engine type")
```
This plot suggests a positive linear trend between horsepower and fuel consumption (indicating that more powerful cars tend to be less fuel-efficient). The relationship exhibits an increasing variability at higher horsepower levels.

To further explore the differences in fuel consumption between engine types, a boxplot was created. 
```{r}
no_electric <- qatarcars %>% 
  filter(enginetype != "Electric")

ggplot(no_electric, aes(x = enginetype, y = economy, fill = enginetype)) +
  geom_boxplot() +
  theme_minimal()
```
Petrol vehicles greater variabiliy in fuel consumption, while there is only a slight difference in the medians of the consumption of hybrid and petrol. 

In order to further explore the different characteristics of the engine types, a scatterplot of horsepower and performance, by engine type was created. 
```{r}
ggplot(qatarcars, aes(x = horsepower, y = performance, fill = enginetype)) +
  geom_point(aes(colour = enginetype))
```
This plot suggests a negative relationship between horsepower and performance, indicating that vehicles with higher horsepower tend to have lower performance values. The relationship is nonlinear: performance decreases rapidly, as horsepower increases. This pattern is explicitly strong in the cases of cars fueled by petrol. 

horsepower and: 
  - origin / make
  - origin -» boxplot

Horsepower and mass  
```{r}
ggplot(qatarcars, aes(x = horsepower, y = mass, fill = enginetype)) +
  geom_point(aes(color = enginetype))

## todo: explanation and interpretation
```
Horsepower and price, by engine type 
```{r}
ggplot(qatarcars, aes(x = price, y = horsepower)) +
  geom_point()

## maybe here: not a geom point? - not the most informative option to visualize 
```


# 4. Hypotheses 

# 5. Statistical Modelling

# 6. Diagnostics

# 7. Conclusion